A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-02-01 doc/large_strain.tex
Changes to MF section

{\color{red}For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.}
%
In order to obtain gradients with respect to the spatial configuration,
{\color{red}
that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem,}
we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the spatial configuration.
Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.

{\color{red}
Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
}

@@

Note that the SIMD vectorization
{\color{red}
in \texttt{deal.II} \cite{dealII90}
}
is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
{\color{red}
The main reason is that the operations are typically the same on all elements.
For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{For the heterogeneous materials considered here,
the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
}.
For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type
and overloads basic arithmetic operations based on compiler intrinsics.
The core of the sum factorization algorithms is provided by the
\textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
}

>>>>

{\color{red}For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.}
%
In order to obtain gradients with respect to the spatial configuration,
{\color{red}
that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem \changeJP{derived with a Lagrangian formulation},}
we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the spatial configuration.
Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.

{\color{red}
Below, we propose three algorithms to implement the tangent operator of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach \changeJP{, each of which have a different level of abstraction, algorithmic complexity, and memory requirements}.
\changeJP{They are introduced in order of lowest memory footprint to largest:}
}

@@

Note that the SIMD vectorization
{\color{red}
in \texttt{deal.II} \cite{dealII90}
}
is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
{\color{red}
The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements\footnote{%
\changeJP{In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
This motivates the caching strategy employed in Algorithm 3, as the pre-caching of the chosen data ensures a context in which SIMD parallelization of subsequent operations is guaranteed.
Algorithms 1 and 2 sacrifice some of this generality for decreased memory requirements.}
}.
For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{%
For the heterogeneous materials considered here,
the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
}.
For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type
and overloads basic arithmetic operations based on compiler intrinsics.
The core of the sum factorization algorithms is provided by the
\textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
}
